Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-15 WSSSPE3/report/wg_main_funding_spec_expert.tex
Section 5.2: formatting. No content change.
Section 5.2: Minor changes.

Research Software Engineers -- those who contribute to science and scholarship
through software development -- are an important part of the team
needed to deliver 21st century research. However, existing academic structures
and systems of funding do not effectively fund and sustain these skills.
The resulting high levels of turnover and inappropriate incentives
are a significant contributing factor to low levels of reliability and
readability observed in scientific software. Moreover, the absence of skilled and experienced 
developers retards progress in key projects, and at times causes important projects to fail completely.

Effective development of software for advanced research requires
that researchers work closely with scientific software developers who understand
the research domain sufficiently to build meaningful software at a reasonable pace. 
This requires a collaborative approach -- where developers who are fully engaged/invested 
in the research context are co-developing software with domain academics.

@@

The solution we envision entails creating an environment where software developers 
are a stable part of the research team. Such an environment mitigates the risk
of losing a key developer at a critical moment in a projects lifetime, and provides
the benefits of building a store of institutional knowledge about specific projects as well as about software 
development for today's research. Our vision is to find a way to promote a University/research institute environment where
software developers are stable components of research project teams.

One strategy to promote stability is implementing a mechanism for developers to obtain academic
credit for software development work. With such a mechanism in place, traditional academic funding
models and career tracks could properly sustain individuals for whom software development is their
primary contribution to research. A contributing factor to the problem with the current academic reward system is the
devastating effect on an academic
publication record resulting from time in industry; such postings often develop exactly the skills that research software
engineers need, yet returns to university positions following an industry role are penalized by the current structures.
Retention of senior developers is hard, because these people are highly in demand by the economy. However, people who have a
PhD in science and enter industry, may desire to return for diverse reasons, and should be welcomed back.

While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
developing dedicated research software engineering roles within the University, and finding stable funding for those individuals is the most promising mechanism for creating a stable software development staff.

Measures of impact and success for research programming groups, as well as for individual research software engineers, will
be required in order to make the case to the University for continued funding. Research software engineers will not be measured by publications, we hope, but by other measures. Middle-author publications are common for RSEs. Most RSEs welcome co-authorship on papers where the PIs think that the contribution deserves it.

@@

It is hard for an individual PI in a university or college to support dedicated research software engineering resources, as
the need for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is necessary to aggregate this work across multiple research groups.

@@

Success, appropriately measured, will help make the case to such funding bodies for further investment. One might expect that metrics such as improved productivity, software adoption rates, and grant success rates would be sufficient arguments in favor of such a model. However, useful measurement of code cleanliness, and the resulting productivity gains, is an unsolved problem in empirical software engineering. To measure ``what did not go wrong'' because of an intervention is particularly hard.

We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
UK, a collective effort to identify the research software engineering community, with individuals clearly stating ``I'm a research software engineer,'' has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution. There remains the problem of how to "sell" the value of this investment to investigators within the university. This is an issue best addressed by the individual organizations that embark on the plan.

@@

The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSEs who do not identify themselves as such at this time. We will collect information as to the organizational models under which these groups function, and how they are funded. For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not. Are these developers paid by the University or through a program supported by research grants/individual PIs? How did they bootstrap the developer track to get this started? How successful is the university in getting investigators to pay for fractional RSEs?  We will author a report describing our findings, should funding be available to conduct the investigation.

>>>>

Research Software Engineers (RSEs) -- those who contribute to science and scholarship
through software development -- are an important part of the team needed to
deliver 21st century research. However, existing academic structures and systems
of funding do not effectively fund and sustain these skills. The resulting high
levels of turnover and inappropriate incentives are significant contributing
factors to low levels of reliability and readability observed in scientific
software. Moreover, the absence of skilled and experienced developers retards
progress in key projects, and at times causes important projects to fail
completely.

Effective development of software for advanced research requires that
researchers work closely with scientific software developers who understand the
research domain sufficiently to build meaningful software at a reasonable pace.
This requires a collaborative approach -- where developers who are fully
engaged or invested in the research context are co-developing software with domain
academics.

@@

The solution we envision entails creating an environment where software
developers are a stable part of a research team. Such an environment mitigates
the risk of losing a key developer at a critical moment in a projects lifetime,
and provides the benefits of building a store of institutional knowledge about
specific projects as well as about software development for today's research.
Our vision is to find a way to promote a university/research institute
environment where software developers are stable components of research project
teams.

One strategy to promote stability is implementing a mechanism for developers to
obtain academic credit for software development work. With such a mechanism in
place, traditional academic funding models and career tracks could properly
sustain individuals for whom software development is their primary contribution
to research. A contributing factor to the problem with the current academic
reward system is the devastating effect on an academic publication record
resulting from time in industry; such postings often develop exactly the skills
that research software engineers need, yet returns to university positions
following an industry role are penalized by the current structures. Retention of
senior developers is hard, because these people are high in demand by the
economy. However, people who have a PhD in science and enter industry, may
desire to return for diverse reasons, and should be welcomed back.

While development of new mechanisms in the current academic reward system is a
worthy aspirational goal, such a dramatic change in this structure does not seem
likely in a time scale relevant to this working group. Accordingly, our working
party sought alternative solutions that may be achievable within the context of
existing academic structures. The group felt that developing dedicated research
software engineering roles within the university, and finding stable funding for
those individuals is the most promising mechanism for creating a stable software
development staff.

Measures of impact and success for research programming groups, as well as for
individual research software engineers, will be required in order to make the
case to the university for continued funding. Research software engineers will
not be measured by publications, we hope, but by other metrics. Middle-author
publications are common for RSEs. Most RSEs welcome co-authorship on papers
where the PIs think that the contribution deserves it.

@@

It is hard for an individual PI in a university or college to support dedicated
research software engineering resources, as the need and funding for these
activities are intermittent within a research cycle. To sustain this capacity,
therefore, it is necessary to aggregate work across multiple research
groups.

@@

Success, appropriately measured, will help make the case to such funding bodies
for further investment. One might expect that metrics such as improved
productivity, software adoption rates, and grant success rates would be
sufficient arguments in favor of such a model. However, useful measurement of
code cleanliness and the resulting productivity gains is an unsolved problem
in empirical software engineering. To measure ``what did not go wrong'' because
of an intervention is particularly hard.

We finally noted that the institutional case for such groups is made easier by
having successful examples to point to. In the UK, a collective effort to
identify the research software engineering community, with individuals clearly
stating ``I am a research software engineer,'' has been important to the
campaign. It will be useful to the global effort to similarly identify emerging
research software organizations, and also, importantly, to identify
longer-running research software groups, which have in some cases had a long
running \emph{sui-generis} existence, but which now can be identified as part of
a wider solution. There remains the problem of how to ``sell'' the value of this
investment to investigators within a university. This is an issue best
addressed by the individual organizations that embark on the plan.

@@

The first step in moving this strategy forward is to gather a list of groups
that self-identify as research software engineering groups, and to reach out to
other organizations to see if there may be a widespread community of RSEs who do
not identify themselves as such at this time. We will collect information about
the organizational models under which these groups function, and how they are
funded. For example, how many research universities currently fund people in the
RSE track, whether they bear the RSE moniker or not. Are these developers
paid by the university or through a program supported by research
grants/individual PIs? How did they bootstrap the developer track to get this
started? How successful is the university in getting investigators to pay for
fractional RSEs? We will author a report describing our findings, should funding
be available to conduct the investigation.
